---
title: "BrowserViz:  A base class providing simple, extensible message passing between 
 your R session and web browser, for interactive data visualization."
author: "Paul Shannon"
package: BrowserViz
date: "`r Sys.Date()`"
output:
   BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{"BrowserViz: support programmatic access to javascript apps running in your web browser"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

BrowserViz provides the basis for, and a very simple working example of, 
interactive R/browser visualization.  Thus two interactive powerful and complementary
environments are linked, creating a powerful hybrid setting for exploratory data analysis.


This work is motivated by our belief that contemporary web browsers, supporting
HTML5 and Canvas, and running increasingly powerful Javascript libraries (for example, [d3](https://d3js.org/))
and [cytoscape.js](http://js.cytoscape.org/) have become the best setting in which to
develop interactive graphics for exploratory data analysis.  We 
predict that web browsers, already powerful and easily
programmed, will steadily improve in rendiring power and interactivity, and thus
remain the optimal setting for interactive R visualization for years to come.

[Shiny](https://shiny.rstudio.com) and [htmlwidgets](https://www.htmlwidgets.org/) are two very popular packages which provide
solutions to this same general problem:  how does one use the power of web browser graphics from R?  
Both of these package create bindings in R to HTML widgets and Javascript objects.
This creates representations for these objects in R: a button, a d3 scatterplot, an
interactive geographic map. The two packages provide elegant support for these bindings
and a clear path to creating more of them.

The principal different between BrowserViz and these worthy packages
is the use of only **loose coupling** of R and the browser.  In slogan
form, our approach can be summarized, ``let R be R, and Javascript be
Javascript''.  BrowserViz links wwo rich programming environments, but
the environments are kept maximally ignorant of each other.  Only
simple JSON messages pass back and forth, and these are at a high
semantic level: no HTML, CSS or Javascript.  Rather than creating
representations of web objects in R, which would be tight-coupling of
the two environments, BrowserViz provides a style of programming in
which


  -  Web objects are (just as in standard widespread web programming) created in HTML, CSS and Javascript
  -  JSON and websockets provide simple explicit message passing between R and the browser
  -  High-level web objects (i.e., an xy plotter, or a network viewer) are created by and manipulated 
    by Javscript functions at the request of high-level R function calls
  -  Web elements can initiate (call back to) R functions
  -  A traditional event-driven architecture is used throughout, in which events are either R function calls or user (keyboard or mouse) actions in the browser. 

BrowserViz provides a very low threshold for those wishing to create R/web browser visualizations.
This base package hides the  complexity of websocket initiation and message passing.
The websocket communication channel is created with a single R function call.
Passing messages and handling responses is similarly simple.  The
intricacy (or simplicity) of the web browser interface is determined by the programmer. A 
vast collection of easily available books, examples, tutorials, and support websites make web browser
programming especially easy to learn.  The BrowserViz approach will be of
interest to any programmer interested in the visualization of data, and proficient in -- or willing
to learn -- both R and Javascript.

Standalone web sites can be created, but the primary intended audience for this package
is the R programmmer exploring and analyzing data in R and using the browser visualization for the 
indispensable benefits it provides.  We hope that many visualization tools will be created.  We provide
a simple x-y plotter (see BrowserVizDemoto illustrate how to write a BrowserViz subclass
application.  The Biocondcutor RCyjs package  is a full-featured visualization tool for network visualization built
upon [cytoscape.js](http://js.cytoscape.org) and the web browser.  

The BrowserViz class, though a base class intended for subclassing,
includes a simple demo which performs a few elementary browser manipulations,
and queries the browser for some simple state (window size, window title,
browser version).  The principal goal of the package is to provide the the
websocket ``plumbing'' along with a standard (simple, open-ended) message
protocol for communicating between the two environments.


# Standard Message Format

Just as the ubiquitous and language-neutral
[websocket](https://en.wikipedia.org/wiki/WebSocket) protocol provides
the BrowserViz communication mechanism, so does [JSON](https://www.w3schools.com/js/js_json_intro.asp)
provide the message notation.  Native data types in R (a named list, for instance)
and Javascript (an object, with key:value pairs) are easily converted
to and from JSON by libraries standard in each language.   R data.frames are a little trickier.

We have adopted a simple, adaptable data structure flexible enough for all of
the uses so far encountered.  In JSON (and Javascript):
```
    {{cmd: "setBrowserWindowTitle",
      status: "request",
      callback:"handleResponse", 
      payload: "BrowserViz Demo"}}
```

Websocket servers both send and receive messages.  Thus a typical BrowserViz event begins
with sending a message from one environment to the other, and often concludes with some sort of a return or
**callback** message.

   - **cmd**: the name of the operation the sender wishes to be performed by the receiver.
   - **status**: might be ``success'', ``failure'', ``error'', ``deferred response''.
   - **callback**: provided by the sender, this specifies the operation which the receiver is 
     to call *in the client* after it (the receiver) completes the operation it was asked to perform.
   - **payload** An open-ended data structure, sometimes empty, as simple as a character string, as complex
              as any conceivable deeply nested list.  

```{r jsonlite,  results='hide'}
  library(jsonlite)
  msg <- toJSON(list(cmd="setBrowserWindowTitle",
                     status="request", 
                     callback="handleResponse",
                     payload="BrowserViz demo"))
``` 

The callback for this request could be empty, which by convention we encode as the empty string.
The calling code, in R, and the receiving code, in Javascript, only need to be consistent.  If the 
caller provides a non-empty *callback*, the Javascript receiver should craft and send a return message
with the canonical four fields specifying *cmd=callback* and any *payload* the caller expects, 
perhaps

```
    {cmd: "handleResponse",
    status: "success",
    callback:"",
    payload: "BrowserViz Demo"}
```

An empty **payload** could also be used, in which case the **success** status of the
return command is the only information returned from Javascript to R.  All decisions of this
sort are left to the programmer.  Often the same person writes the R and the Javascript code
that talk back and forth over the websocket.  If different programmers are involved, then
careful communication and documentation is required, of the expectations, contraints and
*payload( structure.  


# The Simple BrowserViz ``Application''

We predict that the principal use of BrowserViz will be as a base class for other
rich visualzation packages, and that authors of those derived classes will be be able
to proceed without any direct involvement in the nuts and bolts of websocket creation
and handling.  Nonetheless, BrowserViz is a complete R/browser application, albeit
one with only a few features.   These features (R methods on the BrowserViz object), few though
they be, are automatically available to all BrowserViz subclasses.

  - port
  - ready
  - browserResponseReady
  - getBrowserResponse
  - closeWebSocket
  - send
  - getBrowserWindowSize
  - getBrowserWindowTitle
  - setBrowserWindowTitle
  - roundTripTest
  - displayHTMLInDiv

The last two of these methods are used in daily tests of this package.  Conceptually
simple, they also illustrate basic use:


Note that the **browserVizBrowserFile** mentioned in the second and
sixth lines below, and which is included in this package, is a
combination of HTML, javascript and css.  When you construct a
BrowerViz application, you supply this filename; this file is sent to
the browser.  When received and interpreted in the browser, the 
HTML is rendered into the page, and the Javascript code executes,
requesting a websocket connection right back to the R session's
http/websocket server.  After the connection is established, messages
can flow back and forth between the two environments, R and the
web browser.


```{r roundTripTest,  eval=FALSE, results='hide'}
library(BrowserViz)
browserVizBrowserFile <- system.file(package="BrowserViz", "browserCode",
                                     "dist", "bvDemoApp.html")
PORT_RANGE <- 12111:12120
if(BrowserViz::webBrowserAvailableForTesting()){
   bvApp <- BrowserViz(browserFile=browserVizBrowserFile, quiet=TRUE)
   data <- list(lowercase=letters, uppercase=LETTERS)
   json.returned <- roundTripTest(bvApp, data)
   data.returned <- fromJSON(json.returned)
   message(sprintf("    %5d bytes exchanged", nchar(json.returned)))
   stopifnot(data == data.returned)
   html <- sprintf("<h3>round trip of json-encoded data, %d chars</h3>",
                   nchar(json.returned))
   displayHTMLInDiv(bvApp, html, "bvDemoDiv")
   }
```








